---
title: "SIR Demography Example"
author: "Madeline Peters"
date: "31/01/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We'll use a basic SIR model with demography:

$\frac{dS}{dt}=\mu*(S+I+R) + \alpha R-\beta S I - \delta S\\ \\ \frac{dI}{dt}=\beta S I - \kappa I -\delta I\\ \frac{dR}{dt}=\kappa I - \alpha R - \delta R$ 

where $\alpha$ is the rate of immunity loss, $\beta$ is the transmission coefficient, $\kappa$ is the recovery rate, $\mu$ is the birth rate and $\delta$ is the natural mortality rate.

## Load required packages
```{r}
library(tidyverse)
library(deSolve)
library(fda)
library(CollocInfer)
```

## Set seed
```{r}
set.seed(1234)
```

# Generate main data
Define vectors of names for the indices of arrays containing parameters and values of the trajectories:
```{r}
SIRvarnames = c("S","I","R")
SIRparnames = c("alpha","beta","kappa","mu","delta")
```

To use the function "lsoda", we need to specify initial conditions, respectively:
```{r}
x0 = c(10000,1,0)
names(x0) = SIRvarnames
```

Specify the initial values for the parameter vector $\theta$ = c($\alpha$,$\beta$,$\kappa$,$\mu$,$\delta$):
```{r}
SIRpars = c(0.4,0.001,0.5,0.01,0.025)
names(SIRpars) = SIRparnames
```

We need a function to evaluate the SIR equations. The function should return a names list, with a member names fn and containing the function evaluation
```{r}
SIR.ode = function(times,x,p){
  dx = x
  dimnames(dx) = dimnames(x)
  dx["S"] = p["mu"] * (x["S"] + x["I"] + x["R"]) + p["alpha"] * x["R"] - p["beta"] * x["S"] * x["I"] - p["delta"] * x["S"]
  dx["I"] = p["beta"] * x["S"] * x["I"] - p["kappa"] * x["I"] - p["delta"] * x["I"]
  dx["R"] = p["kappa"] * x["I"] - p["alpha"] * x["R"] - p["delta"] * x["R"]
  return(list(dx))
}
```

Now define a time vector from 0 to 20:
```{r}
SIRplottimes = seq(0,20,0.25)
```

Call function `lsoda` to compute the approximate solution at the time values specified above. The output should be an $n \times 5$ matrix where the first column is the evaluation times.
```{r}
out = lsoda(x0,times=SIRplottimes,SIR.ode,SIRpars)
```

Plotting out should show SIR dynamics:
```{r}
out.dat = as.data.frame(out) %>% gather(.,"Type","Count",2:4)
ggplot()+geom_line(data=out.dat,aes(x=time,y=Count,col=Type))+xlab("Time")+ylab("Count")+theme_classic()
```
Now we will generate some data with noise drawn from a normal distribution with mean zero and standard deviation 200. We will ensure all values are positive and turned into counts:
```{r}
SIRtimes = seq(0,20,0.25)
SIRn = length(SIRtimes)
out = lsoda(x0,time=SIRtimes,SIR.ode,SIRpars)
SIRdata = out[,2:4] + 200*matrix(rnorm(3*SIRn), SIRn, 3)
SIRdata[SIRdata < 0] = 0
SIRdata = round(SIRdata)
out.dat2 = as.data.frame(SIRdata) %>% gather(.,"Type","Count",1:3)
out.dat2$Time = SIRtimes
ggplot()+geom_line(data=out.dat2,aes(x=Time,y=Count,col=Type))+xlab("Time")+ylab("Count")+theme_classic()
```
# Assuming recovered $R$ individuals cannot be measured

We're going to pretend we don't have a count of recovered individuals, so that we just have $SE$ and $I$ measurements:
```{r}
SIRdata2 = SIRdata
SIRdata2[,3] = NA
```

## Set up equations for use with *CollocInfer*

For the SIR equations, we now write a function for use with *CollocInfer*:
```{r}
SIR.fun = function(times,x,p,more){
  dx = x
  dx[,"S"] = p["mu"] * (x[,"S"] + x[,"I"] + x[,"R"]) + p["alpha"] * x[,"R"] - p["beta"] * x[,"S"] * x[,"I"] - p["delta"] * x[,"S"]
  dx[,"I"] = p["beta"] * x[,"S"] * x[,"I"] - p["kappa"] * x[,"I"] - p["delta"] * x[,"I"]
  dx[,"R"] = p["kappa"] * x[,"I"] - p["alpha"] * x[,"R"] - p["delta"] * x[,"R"]
  return(dx)
}
```

## Obtain initial coefficient estimates using *fda* package

Set up basis functions:
```{r}
SIRrange = c(0,20)
breaks = seq(0,20,0.25)
SIRbasis = create.bspline.basis(SIRrange,norder=4,breaks=breaks)
```

Obtain initial coefficients for $S$ and $I$ only by penalizing the second derivative and using a lambda (roughness weight) of 10. Set coefficients for the unobserved state variables to zero.
```{r}
fd0 = smooth.basis(SIRtimes, SIRdata2[,1:2], fdPar(SIRbasis, int2Lfd(2), 10))
coef0 = fd0$fd$coef
coefs02 = matrix(0,nrow=nrow(coef0),ncol=3)
rownames(coefs02) = rownames(coef0)
coefs02[,1:2] = coef0[,1:2]
colnames(coefs02) = c("S","I","R")
```

## Make `lik` and `proc` objects
```{r}
par.guess = SIRpars
out = LS.setup(pars=par.guess,coefs=coefs02,basisvals=SIRbasis,fn=SIR.fun,lambda=100,times=SIRtimes)
lik = out$lik
proc = out$proc
```

Obtain updated coefficients by using `FitMatchOpt`:
```{r eval=FALSE, include=FALSE}
Fres3 = FitMatchOpt(coefs02,c(3),par.guess,proc,meth="maxNR")
```

## Proceed with profiling
```{r}
Ores4 = outeropt(SIRdata,SIRtimes,par.guess,Fres3$coefs,lik,proc,out.meth="optim",in.meth="optim")
Ores4$pars #false convergence
```
Obtain confidence intervals on parameter estimtes:
```{r}
covar = Profile.covariance(Ores4$pars, times = SIRtimes, data = SIRdata, coefs = Ores4$coefs, lik = lik, proc = proc)
covar
CIs = cbind(Ores4$pars - 2 * sqrt(diag(covar)),Ores4$pars + 2 * sqrt(diag(covar)) )
rownames(CIs) = SIRparnames
CIs
```


